Computing in the Field of Rationals

### 2012-07-05 fields.tex
_

  This paper connects the line of work starting with
  Filinski\cite{Filinski:1989:DCI:648332.755574} on the duality of
  computation \cite{Curien:2000,DBLP:conf/rta/Wadler05} with work on
  information preserving computation \cite{infeffects,rc2011}.
  Andrzej Filinski's Masters thesis
  \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
  symmetry underlying computation in the \lcal. Filinski showed that
  values are dual to continuations and that functions are dual to
  delimited continuations, thereby suggesting that
  values-and-functions and continuations-and-delimited-continuations
  are mirror images of the same phenomena.

  Previous work on \emph{information effects} established the notion
  of information preserving computation \cite{infeffects} and the core
  calculus for computing with the isomorphisms of finite types,
  {{Pi}}. Given the first-order nature of {{Pi}} however it was
  unclear how one may express first-class functions and duality in the
  sense of Filinski.

  We present a computational model whose types are \emph{the field of
    rational numbers}. This computational model is derived
  systematically from the {{Pi}} by considering a symmetric notion of
  duality. Unlike the \lcal which shows a classical De Morgan duality,
  here we have two axis of dualization and hence two dualities -- an
  additive duality, namely negative types and a multiplicative
  duality, namely fractional types.

@@

Andrzej Filinski's Masters thesis
\cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
underlying computation in the \lcal. Filinski showed that values are
dual to continuations, functions are dual to delimited continuations
and a De Morgan style duality by treating pairs of values as dual to
sums (choice) of continuation and sums of values as dual to pairs of
continuations. He, furthermore, introduced a symmetric extension of
the $\lambda$-calculus in which call-by-value is dual to call-by-name.

Symmetry between the value fragment and the continuation fragment was
refined by Curien and Herbelin \cite{Curien:2000} who introduced the
sequent-style $\mu\tilde{\mu}$-calculus, that exhibits symmetries
between values and continuations and between call-by-value and
call-by-name. The calculus includes the type $A-B$ which is the dual
of implication, i.e., a value of type $A-B$ is a context expecting a
function of type $A \rightarrow B$. Alternatively a value of type
$A-B$ is also explained as a \emph{pair} consisting of a value of type
$A$ and a continuation of type $B$. This is to be contrasted with our
interpretation of a value of that type as \emph{either} a value of $A$
or a demand for a value of type $B$. The duality between call-by-name
and call-by-value was further investigated by Selinger using control
categories~\cite{Selinger:2001:CCD:966910.966911}. The
$\mu\tilde{\mu}$-calculus was further analyzed and extended by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}. The extension gives
no interpretation to the subtraction connective and like the original
symmetric calculus of Filinski, introduces a duality that relates sums
to products and vice-versa.

Previous work on \emph{information effects} established the logically
reversible computational model {{Pi}} (and its extension {{Pi0}})
wherein computation is inherently information preserving. The
computational model is derived from the isomorphisms of finite-types,
every computation admits an adjoint that expresses the logical
``inverse'' of a computation and is complete for combinational
circuits. The term model of {{Pi}} is a so-called commutative
bimonoidal category, i.e. it corresponds to the categorification of a
commutative semiring with two commutative monoidal structures
{{(0,+)}} and {{(*,1)}} with {{*}} distributing over {{+}}.

The language {{Pi}} however lacks a natural representation for
first-class functions due to the first-order nature of its values. In
the light of the work of duality of computation it was felt that
``right model'' for expressing functions in {{Pi}} would also express
symmetries similar to those in the calculi of Filinski, Curien and
Herbelin and Wadler. The technical goal of this paper is to present
such a symmetric dual extension of {{Pi}}.

@@

\begin{definition}[{{eval(c, v)}}]
For {{c:b1<->b2}} and {{v:b1}} we define {{eval(c, v)=v'}} such that
\begin{enumerate}
\item {{v':-b2+b1}} and
\item {{<action(c), right v, []> |-->* <action(c),v',[]>^^^}}
\end{enumerate}
\end{definition}

>>>>

  Previous work on information effects introduced the computational
  model {{Pi}} which lacked a natural notion of first-class functions.
  This paper connects the line of work starting with
  Filinski\cite{Filinski:1989:DCI:648332.755574} on the duality of
  computation \cite{Curien:2000,DBLP:conf/rta/Wadler05} with work on
  information preserving computation \cite{infeffects,rc2011}.

  %% Andrzej Filinski's Masters thesis
  %% \cite{Filinski:1989:DCI:648332.755574} suggested a remarkable
  %% symmetry underlying computation in the \lcal. Filinski showed that
  %% values are dual to continuations and that functions are dual to
  %% delimited continuations, thereby suggesting that
  %% values-and-functions and continuations-and-delimited-continuations
  %% are mirror images of the same phenomena.

  %% Previous work on \emph{information effects} established the notion
  %% of information preserving computation \cite{infeffects} and the core
  %% calculus for computing with the isomorphisms of finite types,
  %% {{Pi}}. Given the first-order nature of {{Pi}} however it was
  %% unclear how one may express first-class functions and duality in the
  %% sense of Filinski.

  We present a computational model whose types are \emph{the field of
    rational numbers}. This computational model is derived
  systematically from the {{Pi}} by considering a symmetric notion of
  duality. Unlike the \lcal which shows a classical De Morgan duality,
  here we have two axis of dualization and hence two dualities -- an
  additive duality, namely negative types and a multiplicative
  duality, namely fractional types. Intuitively, values of negative
  types are values that flow ``backwards'' to satisfy demands and
  values of fractional types are values that impose constraints on
  their context.

@@

Andrzej Filinski's Masters thesis
\cite{Filinski:1989:DCI:648332.755574} suggested a remarkable symmetry
underlying computation in the \lcal. Filinski inroduced a symmetric
extension of the \lcal in which he showed that values are dual to
continuations, functions are dual to delimited continuations, pairs of
values as dual to sums (choice) of continuations
({{(a*b)^bot<->a^bot+b^bot}}), sums of values are dual to pairs of
continuations ({{(a+b)^bot<->a^bot*b^bot}}) and that call-by-value is dual to call-by-name.

The symmetries exposed by Filinski suggested that values/continuations
and functions/delimited continuations are different aspects of the same
computational phenomenon and led to much work in this area.  Symmetry
between the value fragment and the continuation fragment was refined
by Curien and Herbelin \cite{Curien:2000} who introduced the
sequent-style $\mu\tilde{\mu}$-calculus, that exhibits symmetries
between values and continuations and between call-by-value and
call-by-name.  The duality between call-by-name and call-by-value was
further investigated by Selinger using control
categories~\cite{Selinger:2001:CCD:966910.966911} and by
Wadler~\cite{Wadler:2003,DBLP:conf/rta/Wadler05}.

%% The calculus includes the type $A-B$ which is the dual
%% of implication, i.e., a value of type $A-B$ is a context expecting a
%% function of type $A \rightarrow B$. Alternatively a value of type
%% $A-B$ is also explained as a \emph{pair} consisting of a value of type
%% $A$ and a continuation of type $B$. This is to be contrasted with our
%% interpretation of a value of that type as \emph{either} a value of $A$
%% or a demand for a value of type $B$.

%% The extension gives no interpretation to the subtraction connective
%% and like the original symmetric calculus of Filinski, introduces a
%% duality that relates sums to products and vice-versa.

Previous work on \emph{information effects} established the logically
reversible computational model {{Pi}} (and its extension {{Pi0}})
wherein computation is information/entropy preserving. The computational model
is derived from the isomorphisms of finite-types, is complete for
combinational circuits and every computation admits an adjoint which
is the ``inverse of the computation''. The term model of {{Pi}}
corresponds to the categorification of a commutative semiring with two
commutative monoidal structures {{(0,+)}} and {{(*,1)}} with {{*}}
distributing over {{+}} and is sometimes called a commutative bimonoidal
category.

%% The language {{Pi}} however lacks a natural representation for
%% first-class functions due to the first-order nature of its values. In
%% the light of the work of duality of computation it was felt that
%% ``right model'' for expressing functions in {{Pi}} would also express
%% symmetries similar to those in the calculi of Filinski, Curien and
%% Herbelin and Wadler.

The technical goal of this paper is to present the compuational model
{{PiEE}}, a Filinski-style symmetric dual extension of {{Pi}}.

@@

\begin{definition}[{{eval(c, v)}}]
Given {{c:b1<->b2}} and {{v:b1}} we define {{eval(c, v)=v'}} such that
\begin{enumerate}
\item {{right v:-b2+b1}}, {{v':-b2+b1}} and
\item {{<action(c), right v, []> |-->* <action(c),v',[]>^^^}}
\end{enumerate}
\end{definition}
